Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated

### 2024-10-22 analysis.Rmd
changes made to new_manuscript

```{r message=FALSE, warning=FALSE, cache=cache, fig.width=20, fig.height=20}
p_att <- dfsub |>
  select(Participant, starts_with("Att")) |>
  pivot_longer(-Participant) |>
  # mutate(name = str_remove(name, "Cor_")) |>
  ggplot(aes(x = Participant, y = value)) +
  geom_bar(aes(fill = name), stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("black", "#2196F3", "#3F51B5", "#673AB7")) +
  scale_color_manual(values = c("black", "#2196F3", "#3F51B5", "#673AB7")) +
  see::theme_modern() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  labs(y = "Score", fill = "") +
  guides(color = "none") +
  ggside::geom_ysidedensity(aes(x = after_stat(scaled), color = name)) +
  ggside::theme_ggside_void() +
  ggside::scale_ysidex_continuous(expand = c(0, 0))

@@

df <- df |>
  filter(!Participant %in% c(outliers))
```

>>>>

```{r message=FALSE, warning=FALSE, cache=cache, fig.width=20, fig.height=20}

p_att <- dfsub |>
  select(Participant, starts_with("Att")) |>
  tidyr::pivot_longer(-Participant) |>
  # mutate(name = str_remove(name, "Cor_")) |>
  ggplot2::ggplot(aes(x = Participant, y = value)) +
  geom_bar(aes(fill = name), stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("black", "#2196F3", "#3F51B5", "#673AB7", "#843c54","#03989e", "#588163")) +
  scale_color_manual(values = c("black", "#2196F3", "#3F51B5", "#673AB7","#843c54","#03989e", "#588163")) +
  see::theme_modern() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  labs(y = "Score", fill = "") +
  guides(color = "none") +
  ggside::geom_ysidedensity(aes(x = after_stat(scaled), color = name)) +
  ggside::theme_ggside_void() +
  ggside::scale_ysidex_continuous(expand = c(0, 0))

@@

df <- df |>
  filter(!Participant %in% c(outliers))

### 2024-10-22 submission/new_manuscript/manuscript.qmd
changes made to new_manuscript

Advancements in technology have now made it possible to create
near-perfect simulations that are indistinguishable from reality. These
simulations are more affordable, faster to produce, and more accessible
than at any other time in human history. These artificial, yet realistic
constructs permeate all areas of life through immersive works of
fiction, deep fakes (real-like images and videos generated by deep
learning algorithms), virtual and augmented reality (VR and AR),
artificial beings (artificial intelligence "bots" with or without a
physical form), fake news and skewed narratives, of which ground truth
is often hard to access [@nightingale2022]. Such developments not only
carry important consequences for the technological and entertainment
sectors, but also for security and politics - for instance if used for
propaganda and disinformation, recruitment into malevolent
organizations, or religious indoctrination [@pantserev2020]. This issue
is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

Literature shows that the context surrounding a stimulus often plays an
important role in the assessment of its reality [a process henceforth
referred to as *simulation monitoring*, @makowski2018cognitive;
@makowski2019phenomenal]. With the extensive search and processing of
cues within ambiguous stimuli being an increasingly complex and
cognitively effortful strategy [@michael2021source;
@susmann2021persuasion], people tend to draw on peripheral contextual
cues (**Figure 1**), such as the source of the stimulus (e.g., which
journal was the information published in), and its credibility,
authority and expertise, to help facilitate their evaluation
[@petty1986elaboration; @susmann2021persuasion; @michael2021source].
However, the atomization and decontextualization of information allowed
by online social media (where text snippets or video excerpts are often
mass-shared with little context) makes this task progressively difficult
[@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear
contextual information, what drives our beliefs of reality?

@@

AI-generated content, in particular realistic images, is becoming
commonplace and carries important risks for misinformation and
black-mailing [@viola2023designed], emphasizing the need to understand
the different components that come into play in the formation of reality
beliefs. This exploratory study primarily aims at investigating the
effect of facial attractiveness on simulation monitoring, i.e., on the
beliefs that an image is real or artificially generated. Based on the
*affective reality theory* [@makowski2023affective], which suggests that
salient and emotional stimuli are perceived to be more real (up to a
point of reversal after which beliefs of fiction becomes used an emotion
regulation strategy), we hypothesize a quadratic relationship between
perceived realness and attractiveness: faces rated as highly attractive
or unattractive will more likely be believed to be real. We expect a
similar relationship with trustworthiness ratings given its
well-established link with attractiveness [@liefooghe2022faces;
@bartosik2021you; @garrido2017kdef; @little2011facial], and a positive
relationship with familiarity (as more familiar faces would appear as
more salient, self-relevant and anchored in reality). Additionally, we
will further explore the link shared by dispositional traits, such as
personality and attitude towards AI, with simulation monitoring
tendencies. Importantly, this study does investigate the discriminative
accuracy between "true" photos and "true" artificially-generated images
(which we consider more a technological issue than a psychological one),
focusing on the beliefs that a stimulus is real or fake, independently
of its true nature.

@@

All the material (preregistration, experiment demo, experiment code, raw
data, analysis script with complementary figures and analyses, etc.) is
available at
[**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}.

@@

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), **to potentially test and mitigate the potential
effect of expectations/beliefs about AI.** **These items were** mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->
["How attractive would you say you are?", @marcinkowska2021self] and the
other measuring physical attractiveness
<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys.

@@

We excluded 5 participants that either failed 2 (\>= 66.6%) or more
attention check questions, took an implausibly short time to finish the
questionnaires or had incomplete responses. The final sample included
145 participants (Mean age = 28.3, SD = 9.0, range: \[19, 66\]; Sex:
48.3% females, 51.0% males, 0.7% others).

**Data Analysis.** The real-fake ratings (measured originally with a
\[-1, 1\] analog scale) were converted into two scores, corresponding to
two conceptually distinct mechanisms: the dichotomous *belief* (real or
fake, based on the sign of the rating) and the *confidence* (the
rating's absolute value) associated with that belief. The former was
analyzed using logistic mixed models, which modelled the probability of
assigning a face to the real (\>= 0) as opposed to fake (\< 0). The
latter, as well as the other face ratings (attractiveness, beauty,
trustworthiness and familiarity), was modelled using mixed beta
regressions (suited for outcome variables expressed in percentages). The
models included the participants and stimuli as random factors.

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the the two remaining variables *mutatis
mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors.

@@

Several limitations have to be noted. The current experimental paradigm
required participants to judge the realness of faces they had prior
exposure to (which was done to prevent reality judgements from
influencing the other ratings). Although the effect of re-exposure delay
was negligible, the potential bias induced by face familiarity (as
compared to judging completely new items) cannot be discarded. Future
studies could examine that by incorporating novel face images or
increasing the duration of the re-exposure delay.

@@

Finally, it is important to note that although consistent in their
directions across models and variables, the magnitude of the effects
found in the study was relatively small, suggesting that the facial
appraisals measured in the study were not the key determinants of
simulation monitoring. Hence, beyond exploring new potential mechanisms,
future studies should include a more thorough debriefing to try to
capture what conscious strategies (if any) the participants used (e.g.,
focusing on some features of the stimulus - like hair or eyes in the
case of faces) to guide their reality beliefs.

>>>>

Advancements in technology have now made it possible to create
near-perfect simulations that are indistinguishable from reality with an
ease, affordability and accessibility that are unprecedented in Human
history.These artificial, yet realistic constructs permeate all areas of
life through immersive works of fiction, deep fakes (real-like images
and videos generated by deep learning algorithms), virtual and augmented
reality (VR and AR), artificial beings (artificial intelligence "bots"
with or without a physical form), fake news and skewed narratives, of
which ground truth is often hard to access [@nightingale2022]. Such
developments not only carry important consequences for the technological
and entertainment sectors, but also for security and politics - for
instance if used for propaganda and disinformation, recruitment into
malevolent organizations, or religious indoctrination [@pantserev2020].
This issue is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

Literature shows that the context surrounding a stimulfus often plays an
important role in the assessment of its reality [a process henceforth
referred to as *simulation monitoring*, @makowski2018cognitive;
@makowski2019phenomenal]. With the extensive search and processing of
cues within ambiguous stimuli being an increasingly complex and
cognitively effortful strategy [@michael2021source;
@susmann2021persuasion], people tend to draw on peripheral contextual
cues (**Figure 1**), such as the source of the stimulus (e.g., which
journal was the information published in), and its credibility,
authority and expertise, to help facilitate their evaluation
[@petty1986elaboration; @susmann2021persuasion; @michael2021source].
However, the atomization and decontextualization of information allowed
by online social media (where text snippets or video excerpts are often
mass-shared with little context) makes this task progressively difficult
[@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear
contextual information, what drives our beliefs of reality?

@@

AI-generated content, in particular realistic images, is becoming
commonplace and carries important risks for misinformation and
black-mailing [@viola2023designed], emphasizing the need to understand
the different components that come into play in the formation of reality
beliefs. This exploratory study primarily aims at investigating the
effect of facial attractiveness on simulation monitoring, i.e., on the
beliefs that an image is real or artificially generated. Based on the
*affective reality theory* [@makowski2023affective;
@makowski2018cognitive], which suggests that salient and emotional
stimuli are perceived to be more real (up to a point of reversal after
which beliefs of fiction becomes used an emotion regulation strategy),
we hypothesize a quadratic relationship between perceived realness and
attractiveness: faces rated as highly attractive or unattractive will
more likely be believed to be real. We expect a similar relationship
with trustworthiness ratings given its well-established link with
attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef;
@little2011facial], and a positive relationship with familiarity (as
more familiar faces would appear as more salient, self-relevant and
anchored in reality). Additionally, we will further explore the link
shared by dispositional traits, such as personality and attitude towards
AI, with simulation monitoring tendencies. Importantly, this study does
investigate the discriminative accuracy between "true" photos and "true"
artificially-generated images (which we consider more a technological
issue than a psychological one), focusing on the beliefs that a stimulus
is real or fake, independently of its true. **In other words, the
present study investigates the psychological process that leads to
different *beliefs* of reality, rather than the discrimination between
real faces and actual AI-generated ones, which largely depends on the
technological quality of the AI-generation process.**

@@

All the material (preregistration ***[FOOTNOTE: This approach diverges from the preregistration in several key ways. First, the phrasing of items was modified from "Assuming the face you saw was of a real individual, how..." to "I find this person..." A new attractiveness scale (i.e., Beauty) was introduced to capture a more objective measure of attraction. Finally, the data analysis method was altered from Bayesian Mixed Models due to computational limitations, as we were unable to run these models on a high-performance cluster.]***, experiment demo, experiment code, raw
data, analysis script with complementary figures and analyses, etc.) is
available at
[**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}.

@@

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), **to potentially test and mitigate the potential
effect of expectations/beliefs about AI.** **These items were** mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->
["How attractive would you say you are?", @marcinkowska2021self] and the
other measuring physical attractiveness
<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys. **All Cronbach's alpha values were within
the acceptable to excellent range, except for the neuroticism subscale
of the Mini-IPIP6 and the negative subscale of the GAAI, which were
poor, and the Expectations about AI scale, which was questionable (Gliem
& Gliem, 2003; see supplementary material for the details of the
reliability analysis).**

@@

We excluded 5 participants that either failed 2 (\>= 66.6%) or more
attention check questions, took an implausibly short time to finish the
questionnaires or had incomplete responses. Out of the 5 participants
excluded, 2 Participants were excluded because they failed 2 out of 3
attention checks, 1 because they did not answer the sexual orientation
question, which made further analysis impossible, and 2 had an abnormal
low agreement (r \< 0.1) between the beauty and the attractiveness
ratings (possibly indicating random responses as these two scales
exhibited a higher correlation for the other participants).The final
sample included 145 participants (Mean age = 28.3, SD = 9.0, range:
\[19, 66\]; Sex:48.3% females, 51.0% males, 0.7% others).

**Data Analysis.** The real-fake ratings (measured originally with a
\[-1, 1\] analog scale) were converted into two scores, corresponding to
two conceptually distinct mechanisms: the dichotomous *belief* (real or
fake, based on the sign of the rating) and the *confidence* (the
rating's absolute value) associated with that belief. The former was
analyzed using logistic mixed models, which modelled the probability of
assigning a face to the real (\>= 0) as opposed to fake (\< 0). The
latter, as well as the other face ratings (attractiveness, beauty,
trustworthiness and familiarity), was modelled using mixed beta
regressions (suited for outcome variables expressed in percentages). The
models included the participants and stimuli as random **intercepts with
no nested variables**.

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the the two remaining variables *mutatis
mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors. \

@@

Several limitations have to be noted. The current experimental paradigm
required participants to judge the realness of faces they had prior
exposure to (which was done to prevent reality judgements from
influencing the other ratings). Although the effect of re-exposure delay
was negligible, the potential bias induced by face familiarity, that is
by re-presenting the same face stimuli twice, as compared to judging
completely new items, cannot be discarded. Future studies could examine
that by incorporating novel face images or increasing the duration of
the re-exposure delay.

@@

Finally, it is important to note that although consistent in their
directions across models and variables, the magnitude of the effects
found in the study was relatively small, suggesting that the facial
appraisals measured in the study were not the key determinants of
simulation monitoring. Hence, beyond exploring new potential mechanisms,
future studies should include a more thorough debriefing to try to
capture what conscious strategies (if any) the participants used (e.g.,
focusing on some features of the stimulus - like hair or eyes in the
case of faces) to guide their reality beliefs. **Additionally, the role of specific facial features, like perceived dominance, warmth or gender, would be an interesting avenue to explore in future studies, in particular with paradigms directly manipulating these dimensions (for instance using AI to generate faces of different characteristics).**
